A Monte Carlo Synthetic-Acceleration Method for Solving the Thermal Radiation Diffusion Equation

### 2012-08-29 doc/jcomp2012/jcomp2012.tex
Updated paper.

    We present a novel synthetic-acceleration based Monte Carlo method for
    solving the equilibrium thermal radiation diffusion equation in three
    dimensions. The algorithm performance is compared against traditional
    solution techniques using a Marshak benchmark problem and a more complex
    multiple material problem. Our results show that not only can our Monte
    Carlo method be an effective solver for sparse matrix systems, but also
    that it performs competitively with deterministic methods including
    preconditioned conjugate gradient while producing numerically identical
    results. We also discuss various aspects of preconditioning the method and
    its general applicability to broader classes of problems.

@@

In previous work~\cite{evans_2003}, we developed a residual Monte Carlo method
that was an application of Halton's Sequential Monte Carlo
method~\cite{halton_1962,halton_1994} to the one-dimensional equilibrium (1T)
thermal radiation diffusion equation.  At that time, Halton's method was not
widely known in the general computational physics community.  While extending
our method to three dimensions, we have realized that the Sequential Monte
Carlo method is actually a variant of iterative refinement cast as a residual
method. We have developed a new method in which Monte Carlo is used to
accelerate a fixed-point (Richardson) iteration. This algorithm surpasses our
previous one dimensional method and allows for extension to more general
solution techniques.

The purpose of this study is to demonstrate an efficient Monte Carlo solution
method for three dimensional (3D), time-dependent, discrete systems. In this
work, our previous investigations are extended in the following manner:
\begin{itemize}
\item we develop a rigorous synthetic-acceleration Monte Carlo method
  for solving sparse matrix systems;
\item we apply this method to the 1T thermal radiation diffusion
  equation in three dimensions;
\item we compare numerical and performance results of the Monte Carlo
  method against traditional deterministic solution methods.
\end{itemize}

@@

Here we derive the equations required to solve
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) using the MCSA method.  As
discussed in \S~\ref{sec:iter-refin-monte}, the MCSA method requires two
solutions:
\begin{enumerate}
\item A fixed-point iteration to estimate $\bphi^{l+1/2}$.  This is a
  matrix-vector multiply operation.
\item An adjoint Monte Carlo solve to estimate $\delta\bphi^{l+1/2}$.
\end{enumerate}
For clarity, we rewrite the iteration scheme in
Eq.~(\ref{eq:MCSA-iteration}) using the diffusion operator notation
from Eq.~(\ref{eq:preconditioned-diffusion}),
\begin{align*}
   \bphi^{l+1/2} = (\vI - \ve{M}^{-1}\ve{D})\bphi^{l} +
  \ve{M}^{-1}\ve{q}\:, \quad & (\text{fixed-point iteration})\\ 
  \vr^{l+1/2} = \ve{M}^{-1}\ve{q} - \ve{M}^{-1}\ve{D}\bphi^{l+1/2}\:,
  \quad & (\text{compute residual})\\ 
  \ve{M}^{-1}\ve{D}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:, \quad&
  (\text{estimate $\delta\phi$ using adjoint Monte Carlo method})\\ 
  \bphi^{l+1} = \bphi^{l+1/2} + \delta\bphi^{l+1/2}\:. \quad&
  (\text{update}\ \bphi)
\end{align*}
As described in \S~\ref{sec:iter-refin-monte}, the adjoint Monte Carlo method
only approximately inverts $\ve{M}^{-1}\ve{D}$.  We now describe each of these
in turn.

@@

The adjoint Monte Carlo method described in \S~\ref{sec:adjoint-method} is
used to estimate
\begin{equation}
  \ve{M}^{-1}\ve{D}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:.
\end{equation}
As noted above, the residual acts as the source for this simulation, and the
operator $\ve{M}^{-1}\ve{D}$ is implicitly defined in
Eq.~(\ref{eq:preconditioned-discrete-diffusion}).  There are two Monte Carlo
interpretations that can be applied to this system.  The first follows the
mathematical presentation given in \S~\ref{sec:monte-carlo-matrix}.  Namely,
we form probabilities from the iteration matrix and perform random walks to
generate unbiased estimates of the solution vector, $\delta\bphi^{l+1/2}$,
using the estimator in Eq.~(\ref{eq:adjoint-tally}).

@@

The connectivity of an arbitrary cell is illustrated in the 3D mesh segment in
Fig.~\ref{fig:mesh_segment}.  For a particle residing in the blue cell, any
collision results in a scatter to one of the six adjacent cells since
$p_{ii}=0$ in the preconditioned system.  We can easily calculate the
probabilities for each scattering event from
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) without explicitly forming
the iteration matrix.  In this mesh cell, the probabilities for leakage out of
each face are derived from the equations from each adjacent cell.  Using
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) we can explicitly write the
equations from each adjacent cell, here only noting the contribution from cell
7:
\begin{align*}
  \phi_1 &= \frac{\sigm_{7\rightarrow
      1}}{\sigma_{T\,1}}\frac{\Delta_7}{\Delta_1}\phi_7 +
  \ldots\:,\\ \phi_2 &= \frac{\sigp_{7\rightarrow 2}}
          {\sigma_{T\,2}}\frac{\Delta_7}{\Delta_2}\phi_7 + \ldots\:,\\ 
      \phi_3 &= \frac{\sigm_{7\rightarrow
          3}}{\sigma_{T\,3}}\frac{\Delta_7}{\Delta_3}\phi_7 +
      \ldots\:,\\  \phi_4 &= \frac{\sigp_{7\rightarrow
      4}}{\sigma_{T\,4}}\frac{\Delta_7}{\Delta_4}\phi_7 +
          \ldots\:,\\  \phi_5 &= \frac{\sigm_{7\rightarrow
          5}}{\sigma_{T\,5}}\frac{\Delta_7}{\Delta_5}\phi_7 +
              \ldots\:,\\  \phi_6 &= \frac{\sigp_{7\rightarrow
              6}}{\sigma_{T\,6}}\frac{\Delta_7}{\Delta_6}\phi_7 +
                  \ldots\:,\\ 
\end{align*} 
A quick view of these equations reveals that the only non-zero entries in
column 7 of the iteration matrix are
\begin{equation*}
  \begin{pmatrix}
    h_{1,7} & h_{2,7} & h_{3,7} & h_{4,7} & h_{5,7} & h_{6,7}
  \end{pmatrix}^{T}
  =
  \begin{pmatrix}
    \frac{\sigm_{7\rightarrow
        1}}{\sigma_{T\,1}}\frac{\Delta_7}{\Delta_1}
    \\ \frac{\sigp_{7\rightarrow
        2}}{\sigma_{T\,2}}\frac{\Delta_7}{\Delta_2}
    \\ \frac{\sigm_{7\rightarrow
        3}}{\sigma_{T\,3}}\frac{\Delta_7}{\Delta_3}
    \\ \frac{\sigp_{7\rightarrow
        4}}{\sigma_{T\,4}}\frac{\Delta_7}{\Delta_4}
    \\ \frac{\sigm_{7\rightarrow
        5}}{\sigma_{T\,5}}\frac{\Delta_7}{\Delta_5}
    \\ \frac{\sigp_{7\rightarrow
        6}}{\sigma_{T\,6}}\frac{\Delta_7}{\Delta_6}
  \end{pmatrix}\:,
\end{equation*}
where, following Eq.~(\ref{eq:point_iteration}), $\ve{H} = (\vI -
\ve{M}^{-1}\ve{D})$ is the iteration matrix.  Now, using
Eq.~(\ref{eq:adjoint-probability}), we define the scattering probabilities for
cell 7 as
\begin{equation*}
  P_{7\rightarrow n} = p_{7,n} = \frac{|h_{n,7}|}{|h_{1,7} + h_{2,7} +
    h_{3,7} + h_{4,7} + h_{5,7} + h_{6,7}|}\:,\quad n\in
  \{1,\ldots,6\}\:.
\end{equation*}
The weight change for each scattering event is calculated using
Eq.~(\ref{eq:adjoint-weight}),
\begin{equation*}
  W_{m} = \frac{h_{n,7}}{p_{7,n}}W_{m-1}\:,\quad n\in
  \{1,\ldots,6\}\:,
\end{equation*}
where $m$ is the step-index in the particle history.  We also note that, for
the 3D Cartesian mesh considered in this work, the maximum number of non-zero
entries in columns of the iteration matrix is 6, generating a sparse system.
Cells along the problem boundary can have fewer than 6 entries.
\begin{figure}[h!]
  \centerline{ \input{3D_mesh_segment.pdftex_t}}
  \caption{3D mesh-segment.  The particle's current cell (in blue) is
    connected to 6 adjacent cells.  The cell numbers are plotted
    alongside the mesh such that the blue cell has index 7, the high
    $z$ cell has index 2, etc.}
  \label{fig:mesh_segment}
\end{figure}

@@

Here we examine the effectiveness of the MCSA method and compare it to
traditional deterministic solution methods.  The deterministic methods
considered are preconditioned Conjugate Gradient (PCG) and preconditioned
fixed-point (Richardson) iteration (PFIX).  The PCG method uses right-left
preconditioning with $\ve{M}=\ve{L}\ve{L}^{-1}$.  The PFIX scheme is used to
solve Eq.~(\ref{eq:preconditioned-diffusion}), which is identical to the
system being solved with MCSA.  We examine each method on two 3D problems: a
Marshak wave problem and a multi-material duct problem.

@@

Figure~\ref{fig:marshak_10} shows the analytic Marshak Wave solution compared
to the MCSA solution.  As opposed to most Monte Carlo techniques, the MCSA
method generates solutions that are numerically equivalent to the PCG and PFIX
results.  Because the MCSA method converges the residual of the system, the
numerical accuracy is determined by the stopping criterion shown in
Eq.~(\ref{eq:stopping-criteria}).  Also, the MCSA method is not bound by the
Central Limit Theorem so convergence is not restricted to
$1/\sqrt{N}$~\cite{halton_1994,evans_2003}.
\begin{figure}[h!]
  \centerline{ \includegraphics[width=5in,clip]{marshak_10.pdf}}
  \caption{ Solution of the Marshak wave problem at $t=100$~ns.  The
    analytic solution is compared to the MCSA solution.  Note that the
    MCSA, PCG, and PFIXED methods all give numerically identical
    answers.}
  \label{fig:marshak_10}
\end{figure}

>>>>

    We present a novel synthetic-acceleration based Monte Carlo method for
    solving the equilibrium thermal radiation diffusion equation in three
    spatial dimensions. The algorithm performance is compared against
    traditional solution techniques using a Marshak benchmark problem and a
    more complex multiple material problem. Our results show that our Monte
    Carlo method is an effective solver for sparse matrix systems. For
    solutions converged to the same tolerance, it performs competitively with
    deterministic methods including preconditioned conjugate gradient. We also
    discuss various aspects of preconditioning the method and its general
    applicability to broader classes of problems.

@@

In previous work~\cite{evans_2003}, we developed a residual Monte Carlo method
that was an application of Halton's Sequential Monte Carlo
method~\cite{halton_1962,halton_1994} to the one-dimensional equilibrium
thermal radiation diffusion equation.  At that time, Halton's method was not
widely known in the general computational physics community.  While extending
our method to three spatial dimensions, we have realized that the Sequential
Monte Carlo method is actually a variant of iterative refinement cast as a
residual method. We have developed a new method in which Monte Carlo is used
to accelerate a fixed-point (Richardson) iteration. This algorithm surpasses
our previous one dimensional method and allows for extension to more general
solution techniques.

The purpose of this study is to demonstrate an efficient Monte Carlo solution
method for three dimensional (3D), time-dependent, discrete systems. In this
work, our previous investigations are extended in the following manner:
\begin{itemize}
\item we develop a rigorous synthetic-acceleration Monte Carlo method
  for solving sparse matrix systems;
\item we apply this method to the one-temperature (1T) thermal radiation
  diffusion equation in three dimensions;
\item we compare numerical and performance results of the Monte Carlo
  method against traditional deterministic solution methods.
\end{itemize}

@@

Here we derive the equations required to solve
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) using the MCSA method.  As
discussed in \S~\ref{sec:iter-refin-monte}, the MCSA method requires two
solutions:
\begin{enumerate}
\item A fixed-point iteration to estimate $\bphi^{l+1/2}$.  This is a
  matrix-vector multiply operation.
\item An adjoint Monte Carlo solve to estimate $\delta\bphi^{l+1/2}$.
\end{enumerate}
For clarity, we rewrite the iteration scheme in Eq.~(\ref{eq:MCSA-iteration})
using the diffusion operator notation from
Eq.~(\ref{eq:preconditioned-diffusion}),
\begin{align*}
   \bphi^{l+1/2} = (\vI - \ve{M}^{-1}\ve{D})\bphi^{l} +
  \ve{M}^{-1}\ve{q}\:, \quad & (\text{fixed-point iteration})\\
  %%
  \vr^{l+1/2} = \ve{M}^{-1}\ve{q} - \ve{M}^{-1}\ve{D}\bphi^{l+1/2}\:,
  \quad & (\text{compute residual})\\ 
  %%
  \ve{M}^{-1}\hat{\ve{D}}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:, \quad&
  (\text{estimate $\delta\phi$ using adjoint Monte Carlo method})\\
  %%
  \bphi^{l+1} = \bphi^{l+1/2} + \delta\bphi^{l+1/2}\:. \quad&
  (\text{update}\ \bphi)
\end{align*}
As described in \S~\ref{sec:iter-refin-monte}, the adjoint Monte Carlo method
only approximately inverts $\ve{M}^{-1}\hat{\ve{D}}$ as indicated by the
$\hat{\cdot}$ notation.  We now describe each of these in turn.

@@

The adjoint Monte Carlo method described in \S~\ref{sec:adjoint-method} is
used to estimate
\begin{equation}
  \ve{M}^{-1}\hat{\ve{D}}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:.
\end{equation}
As noted above, the residual acts as the source for this simulation, and the
operator $\ve{M}^{-1}\hat{\ve{D}}$ is implicitly defined in
Eq.~(\ref{eq:preconditioned-discrete-diffusion}).  There are two Monte Carlo
interpretations that can be applied to this system.  The first follows the
mathematical presentation given in \S~\ref{sec:monte-carlo-matrix}.  Namely,
we form probabilities from the iteration matrix and perform random walks to
generate unbiased estimates of the solution vector, $\delta\bphi^{l+1/2}$,
using the estimator in Eq.~(\ref{eq:adjoint-tally}).

@@

The connectivity of an arbitrary cell is illustrated in the 3D mesh segment in
Fig.~\ref{fig:mesh_segment}.  For a particle residing in the blue cell, any
collision results in a scatter to one of the six adjacent cells since
$p_{ii}=0$ in the preconditioned system.  We can easily calculate the
probabilities for each scattering event from
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) without explicitly forming
the iteration matrix.  In this mesh cell, the probabilities for leakage out of
each face are derived from the equations from each adjacent cell.  Using
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) we can explicitly write the
equations from each adjacent cell, here only noting the contribution from cell
7:
\begin{align*}
  \phi_1 &= \frac{\sigm_{7\rightarrow 1}}{\sigma_{T\,1}}
  \frac{\Delta_7}{\Delta_1}\phi_7 + \ldots\:,\\ 
  %%
  \phi_2 &= \frac{\sigp_{7\rightarrow 2}}
  {\sigma_{T\,2}}\frac{\Delta_7}{\Delta_2}\phi_7 + \ldots\:,\\
  %%
  \phi_3 &= \frac{\sigm_{7\rightarrow 3}}{\sigma_{T\,3}}
  \frac{\Delta_7}{\Delta_3}\phi_7 + \ldots\:,\\ 
  %%
  \phi_4 &= \frac{\sigp_{7\rightarrow 4}}{\sigma_{T\,4}}
  \frac{\Delta_7}{\Delta_4}\phi_7 + \ldots\:,\\
  %%
  \phi_5 &= \frac{\sigm_{7\rightarrow 5}}{\sigma_{T\,5}}
  \frac{\Delta_7}{\Delta_5}\phi_7 + \ldots\:,\\
  %%
  \phi_6 &= \frac{\sigp_{7\rightarrow 6}}{\sigma_{T\,6}}
  \frac{\Delta_7}{\Delta_6}\phi_7 + \ldots\:,\\ 
\end{align*} 
A quick view of these equations reveals that the only non-zero entries in
column 7 of the iteration matrix are
\begin{equation*}
  \begin{pmatrix}
    h_{1,7} & h_{2,7} & h_{3,7} & h_{4,7} & h_{5,7} & h_{6,7}
  \end{pmatrix}^{T}
  =
  \begin{pmatrix}
    \frac{\sigm_{7\rightarrow
        1}}{\sigma_{T\,1}}\frac{\Delta_7}{\Delta_1}
    \\ \frac{\sigp_{7\rightarrow
        2}}{\sigma_{T\,2}}\frac{\Delta_7}{\Delta_2}
    \\ \frac{\sigm_{7\rightarrow
        3}}{\sigma_{T\,3}}\frac{\Delta_7}{\Delta_3}
    \\ \frac{\sigp_{7\rightarrow
        4}}{\sigma_{T\,4}}\frac{\Delta_7}{\Delta_4}
    \\ \frac{\sigm_{7\rightarrow
        5}}{\sigma_{T\,5}}\frac{\Delta_7}{\Delta_5}
    \\ \frac{\sigp_{7\rightarrow
        6}}{\sigma_{T\,6}}\frac{\Delta_7}{\Delta_6}
  \end{pmatrix}\:,
\end{equation*}
where, following Eq.~(\ref{eq:point_iteration}), $\ve{H} = (\vI -
\ve{M}^{-1}\ve{D})$ is the iteration matrix.  Now, using
Eq.~(\ref{eq:adjoint-probability}), we define the scattering probabilities for
cell 7 as
\begin{equation*}
  P_{7\rightarrow n} = p_{7,n} = \frac{|h_{n,7}|}{|h_{1,7} + h_{2,7} +
    h_{3,7} + h_{4,7} + h_{5,7} + h_{6,7}|}\:,\quad n\in
  \{1,\ldots,6\}\:.
\end{equation*}
The weight change for each scattering event is calculated using
Eq.~(\ref{eq:adjoint-weight}),
\begin{equation*}
  W_{m} = \frac{h_{n,7}}{p_{7,n}}W_{m-1}\:,\quad n\in
  \{1,\ldots,6\}\:,
\end{equation*}
where $m$ is the step-index in the particle history.  We also note that, for
the 3D Cartesian mesh considered in this work, the maximum number of non-zero
entries in columns of the iteration matrix is 6, generating a sparse system.
Cells along the problem boundary can have fewer than 6 entries.
\begin{figure}[h!]
  \centerline{ \input{3D_mesh_segment.pdftex_t}}
  \caption{3D mesh-segment.  The particle's current cell (in blue) is
    connected to 6 adjacent cells.  The cell numbers are plotted
    alongside the mesh such that the blue cell has index 7, the high
    $z$ cell has index 2, etc.}
  \label{fig:mesh_segment}
\end{figure}

@@

Here we examine the effectiveness of the MCSA method and compare it to
traditional deterministic solution methods.  The deterministic methods
considered are preconditioned Conjugate Gradient (PCG) and preconditioned
fixed-point (Richardson) iteration (PFIX).  The PCG method uses right-left
preconditioning with $\ve{M}=\ve{L}\ve{L}^{-1}$, where $\ve{L} = \ve{L}^{T} =
\sqrt{\ve{M}}$.  Naturally, more powerful preconditioning methods could be
used; however, this provides a fair comparison with the preconditioning used
in the MCSA method.  Different precondioning strategies within MCSA will be
the focus of a future paper.  The PFIX scheme is used to solve
Eq.~(\ref{eq:preconditioned-diffusion}), which is identical to the system
being solved with MCSA.  We examine each method on two 3D problems: a Marshak
wave problem and a multi-material duct problem.

@@

Figure~\ref{fig:marshak_10} shows the analytic Marshak Wave solution compared
to the MCSA solution.  As opposed to most Monte Carlo techniques, the MCSA
method generates solutions that are numerically equivalent to the PCG and PFIX
results.  Because the MCSA method converges the residual of the system, the
numerical accuracy is determined by the stopping criterion shown in
Eq.~(\ref{eq:stopping-criteria}).  Also, the MCSA method is not bound by the
Central Limit Theorem so convergence is not restricted to
$1/\sqrt{N}$~\cite{halton_1994,evans_2003}.
\begin{figure}[h!]
  \centerline{ \includegraphics[width=5in,clip]{marshak_10.pdf}}
  \caption{ Solution of the Marshak wave problem at $t=100$~ns.  The analytic
    solution is compared to the MCSA solution.  The MCSA, PCG, and PFIXED
    methods are all converged to the same tolerance and therefore give
    numerically equivalent answers.}
  \label{fig:marshak_10}
\end{figure}
